Natural Selection Across Three Generations of Americans

### 2024-02-23 hrs-selection.qmd
Move & rewrite tests for interaction effects with top coefficients. Test EA and ADHD. Add a figure showing the results.

```{r}
#| label: Top PGS
#| eval: true

for (eth in "white") {
  r_eth <- rand |> subset(ethnicity == eth)

  # the colon interaction estimates the effect separately within each group
  form_pgs_x_ed <- paste0(pgs, "*edu") |>
    map(\(x) reformulate(c(x, "edu"), response = "rlrs"))
  mod_pgs_x_ed[[eth]]  <-
    map(form_pgs_x_ed, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_ed[[eth]] <-
    map(mod_pgs_x_ed[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_ed[[eth]]) <- pgs
  tidy_pgs_x_ed[[eth]] <-
    list_rbind(tidy_pgs_x_ed[[eth]], names_to = "pgs") |>
    filter(pgs == "EA") |>
    mutate(edu = str_remove(term, ".*:"))

  form_pgs_x_inc <- paste0(pgs, "*income_med") |>
    map(\(x) reformulate(c(x, "income_med"), response = "rlrs"))
  mod_pgs_x_inc[[eth]]  <-
    map(form_pgs_x_inc, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_inc[[eth]] <-
    map(mod_pgs_x_inc[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_inc[[eth]]) <- pgs
  tidy_pgs_x_inc[[eth]] <-
    list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
    filter(pgs == "EA") |>
    mutate(income_med = str_remove(term, ".*:income_med"))

  form_pgs_x_mar <- paste0(pgs, "*married") |>
    map(\(x) reformulate(c(x, "married"), response = "rlrs"))
  mod_pgs_x_mar[[eth]]  <-
    map(form_pgs_x_mar, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_mar[[eth]] <-
    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_mar[[eth]]) <- pgs
  tidy_pgs_x_mar[[eth]] <-
    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
    filter(pgs == "EA") |>
    mutate(
      married = str_remove(term, ".*:married")
    )

  form_pgs_x_agefb <- paste0(pgs, "*agefb") |>
    map(\(x) reformulate(c(x, "agefb"), response = "rlrs"))
  mod_pgs_x_agefb[[eth]]  <-
    map(form_pgs_x_agefb, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_agefb[[eth]] <-
    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
    setNames(pgs) |> 
    list_rbind(names_to = "pgs") |>
    filter(pgs == "EA") |>
    mutate(
      agefb = str_remove(term, ".*:agefb")
    ) 
}

eff_ea_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"estimate"]
se_ea_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"std.error"]
p_ea_ed <-  ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))
eff_ea_int_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"estimate"]
se_ea_int_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"std.error"]
p_ea_int_ed <-  ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"p.value"] < 0.05, "*", "")))

eff_ea_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"estimate"]
se_ea_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"std.error"]
p_ea_inc <-  ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))

eff_ea_int_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"estimate"]
se_ea_int_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"std.error"]
p_ea_int_inc <-  ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"p.value"] < 0.05, "*", "")))

eff_ea_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"estimate"]
se_ea_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"std.error"]
p_ea_mar <-  ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))

eff_ea_int_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"estimate"]
se_ea_int_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"std.error"]
p_ea_int_mar <-  ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"p.value"] < 0.05, "*", "")))

eff_ea_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"estimate"]
se_ea_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"std.error"]
p_ea_agefb <-  ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))

eff_ea_int_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"estimate"]
se_ea_int_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"std.error"]
p_ea_int_agefb <-  ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"p.value"] < 0.05, "*", "")))

@@

@tbl-x-bootstraps shows estimates and bootstrapped 95% confidence intervals for the number of scores where the $\delta$ term is positive. Confidence intervals exclude 34 only for income.

@@

>>>>

@@

@tbl-x-bootstraps shows estimates and bootstrapped 95% confidence intervals for the number of scores where the $\delta$ term is positive. Confidence intervals exclude 34 only for income and marriage.

@@

To complement the above cross-PGS analysis, we focus on two scores which we a priori expect to show intergroup differences: educational attainment and ADHD. Both of these showed large differences in @hugh2022human, and have significant main effects in this data. Figure @fig-top-pgs-interactions displays coefficients of these scores on RLRS for each of the different groups. All differences are in the expected direction, with coefficients being larger for the disadvantaged group, except for low- vs- high-education and EA. However, the only significant difference is between the low- and high-income groups for the coefficient of EA, and this does not survive Bonferroni correction. [XXX or we just say "nothing survives B. C.]

```{r}
#| label: stat-top-pgs-interactions

run_interaction_model <- function (pgs, interaction) {
  # This formulation gives you separate estimates of `pgs`
  # for each value of  `interaction`, as well as the main effect
  # for `interaction` itself. We can then test separately
  # for the interaction term using e.g. coeftest()
  # I prefer doing this because the coefficients are easy to interpret
  cross_terms <- glue("{interaction} + {interaction}:{pgs}")
  form_interaction <- reformulate(cross_terms, response = "rlrs")
  rand_white <- rand |> subset(ethnicity == "white")
  mod_interaction <- svyglm_quiet(form_interaction, rand_white)

  mod_interaction
}

test_interaction_term <- function (mod) {
  # All our models have two interaction terms. This finds them
  # and tests if they are equal, using car::linearHypothesis.
  x_terms <- matchCoefs(mod, ":") 
  x_hypothesis <- glue("{x_terms[1]} = {x_terms[2]}")
  car::linearHypothesis(mod, x_hypothesis)
}

interactions <- c(
    Education = "edu", 
    Income    = "income_med", 
    Marriage  = "married",
    AFB       = "agefb"
  )

top_pgs <- c("EA", "ADHD")

regs_to_run <- expand_grid(pgs = top_pgs, interaction = interactions)

# This creates a tibble of regression results
tidy_regs <- regs_to_run |> 
  # Using rowwise() like this preserves the variables pgs and interaction,
  # but makes it easier to refer to variables in list-columns.
  rowwise(pgs, interaction) |> 
  mutate(
    mod = list(run_interaction_model(pgs = pgs, interaction = interaction)),
    tidy = list(broom::tidy(mod, conf.int = TRUE)),
    # A test that the two interaction terms
    # have the same coefficient:
    test = list(test_interaction_term(mod)),
    # P value of the test:
    test.p.value = suppressWarnings(broom::tidy(test)) |> pull(p.value)
  )

# this gives all the coefficients for the models
tidy_coefs <- tidy_regs |> 
  unnest(tidy) |> 
  ungroup()

```{r}
#| label: fig-top-pgs-interactions
#| fig-cap: |
#|   Coefficients of RLRS across different groups for EA and ADHD 
#|   polygenic scores. Lines show 95 per cent confidence intervals. Groups
#|   are: 0-12 years education vs. 13-17 years; below vs. above median
#|   income; all others vs. married; below vs. above median age at first
#|   birth.

pos_dod <- position_dodge(width = 0.2)

tidy_coefs |> 
  # Select only the interaction terms:
  filter(grepl(":", term)) |> 
  mutate(.by = c(pgs, interaction),
    PGS = pgs,
    # gets the pretty name of the group
    Split = names(interactions)[match(interaction, interactions)],
    Group = c("Unprivileged", "Privileged"),
    Difference = ifelse(test.p.value < 0.05, "Difference p < 0.05",
                        "Difference n.s.")
  ) |> 
  ggplot(aes(estimate, Split, colour = Group)) + 
    geom_linerange(aes(xmin = conf.low, xmax = conf.high), alpha = 0.5,
                   position = pos_dod) +
    geom_point(aes(shape = Group, colour = Group, fill = Difference),
               position = pos_dod, 
               shape = "circle filled", size = 2) +
    facet_wrap(vars(PGS), ncol = 1) +
    scale_fill_manual(values = c("n.s." = "white", 
                                 "p < 0.05" = "orange"),
                                 guide = guide_legend(title = NULL)) +
    scale_color_manual(values = c("Privileged" = "steelblue4", 
                                  "Unprivileged" = "red4")) +
    theme_minimal() +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor.x = element_blank()
    ) + 
    labs(x = "Coefficient on RLRS", y = "")
